### 2015-12-01 paper.tex
Add Budi's edits and comments

Sandia National Laboratories, in collaboration with the US Department of Energy
developed the Reference Model devices to provide modeling targets for advancing
marine hydrokinetic (MHK) energy technology \cite{Neary2014}. The Reference
Model 2 (RM2), a cross-flow turbine intended for river environments, was
initially designed by Barone et al. \cite{Barone2011} using the CACTUS vortex
line code \cite{Murray2011}. A scaled physical model of the RM2 was then
designed and built by the Saint Anthony Falls Laboratory (SAFL) to validate the
CACTUS predictions. Experimental performance measurements were significantly
lower than the numerical results \cite{Hill2014}, which was assumed to be a 
result of the physical model's low Reynolds number
\begin{equation}
Re_L = \frac{UL}{\nu},
\label{eq:Re}
\end{equation}
where $U$ and $L$ are characteristic length scales, and $\nu$ is the fluid
kinematic viscosity.

As a cross-flow turbine, the RM2 blades experience large variations in angles of
attack as they rotate about their axis (``cross-flow turbine'' means the axis of
rotation is perpendicular to flow direction---either vertically or
horizontally). These variations become larger as the tip speed ratio
\begin{equation}
\lambda=\frac{\omega R}{U_\infty},
\end{equation}
where $\omega$ and $R$ are the rotor angular velocity and radius, decreases
\cite{Para2002}. Typically, the maximum angles of attack are sufficiently large
that the blades operate under dynamic stall, which is a complex unsteady process
and deviates significantly from static foil behavior, during part of the
rotation. Furthermore, the higher the solidity of a cross-flow turbine, the
lower the optimal tip-speed ratio at which it operates. Marine Hydrokinetic
(MHK) cross-flow turbines typically have higher solidity than cross-flow wind
turbines (VAWT), and hence operate at lower tip-speed ratios. Since MHK turbines
operate in a higher density fluid, unsteady dynamic effects related to the
blades' pitching motion and flow curvature also become more important when
compared to wind turbines.

@@

The need for experimental data that is relevant to full-scale behavior stems
from the need to validate numerical models---most importantly mid-fidelity
models---desirable for MHK developers to predict the performance of their
cross-flow turbine designs, since physical modeling at appropriate scales can be
prohibitively expensive in the early stages of engineering. Furthermore,
Navier--Stokes-based computational fluid dynamics (CFD) simulations require
modeling in 3 dimensions, which generally necessitates high performance
computing---a resource that is not commonly available, and more expensive.

@@

A 1:6 scale model of the RM2 rotor was designed and built. Turbine geometric
parameters were scaled from the RM2 ``rev 0'' design report \cite{Barone2011},
with the exception of the shaft diameter, which was scaled from the SAFL RM2
model \cite{Hill2014}. The hubs were designed to be similar to the SAFL model,
which may aid in comparison of the results, though this was not a top priority.
Geometric parameters are shown in Tab.~\ref{tab:turb-geom} and a drawing of the
turbine design is shown in Fig.~\ref{fig:turbine-drawing}. The turbine model
components---blades, struts, shaft, and center hub sections---were fabricated
from 6061-T6 aluminum, which was hardcoat anodized per MIL-8625-A, type III,
class 2 specifications. CAD models of the turbine are available from
\cite{Bachant2015-RM2-CAD}.

@@

Experiments were performed in the UNH tow/wave tank, a 36 m long facility with a
3.66 m wide by 2.44 m deep cross-section, capable of tow speeds up to 3 m/s,
pictured in Figure~\ref{fig:tow-tank}. The turbine was mounted in a frame built
from NACA 0020 struts, attached to the tow carriage by four linear bearings,
which transfer all streamwise force to a pair of S-beam load cells. The turbine
shaft RPM was controlled by a servo motor system, which allowed prescription of
the turbine tip speed ratio. The load torque was measured by an inline rotary
torque transducer and a load cell mounted at a fixed distance from the servo
motor, providing a redundant measurement. Turbine shaft angle was measured using
the servo drive's emulated encoder output, set to $10^5$ counts per turbine
shaft revolution. Carriage speed, and therefore inflow velocity was measured
using a linear encoder with 10 $\mu$m resolution. All of these
performance-related quantities were sampled as 2 kHz, while the tow tank's
motion controller provided redundant measurements of the carriage speed and
turbine angular velocity sampled at 1 kHz. Turbine wake measurements at 1 meter
downstream were measured with a Nortek Vectrino+ acoustic Doppler velocimeter,
sampling at 200 Hz. A photo and drawing of the experimental setup are shown in
Fig.~\ref{fig:exp-setup-photo} and Fig.~\ref{fig:exp-setup}, respectively.

@@

    \caption{{\bf Experimental setup.} Illustration of the experimental setup.}

    \label{fig:exp-setup}
\end{figure}

@@

\paragraph{Tare drag and torque compensation.} Tare torque and drag runs were
performed to measure the shaft bearing friction torque and turbine mounting
frame drag, respectively. Tare drag runs were performed for each tow speed in
the experiment, for which the mean value was used in data processing. Tare
torque runs were performed by rotating the turbine shaft (without blades) in air
at constant angular velocity for a specified duration, over the range of angular
velocities used throughout the experiment. Tare torque was fit with a linear
regression versus shaft angular velocity, and added to the measured turbine
torque in post-processing.

@@

\begin{itemize}

@@

For each tow, a relevant quasi-steady duration was selected, and then truncated
to include a whole number of blade passages. Over this duration relevant
statistics are calculated.

@@

Mean rotor power and drag coefficients for multiple Reynolds numbers are plotted
versus tip speed ratio in Fig.~\ref{fig:cp-curves} and Fig.~\ref{fig:cd-curves},
respectively. $C_P$ is shown to increase in general with $Re$, along with a
reduction in the optimal tip speed ratio, due to stall delay. These effect
diminish with increasing $Re$, which is expected as the blade boundary layers
transition from laminar to turbulent \cite{Lissaman1983,
Bachant2015-RVAT-Re-dep}. The rotor drag coefficient curves show similar
effects, but vary much less dramatically from each other.

\begin{figure}[h]
\includegraphics[width=\textwidth]{figures/cp_curves}
\caption{{\bf Power Coefficient Curves.}
Mean rotor power coefficient plotted versus mean tip speed ratio for multiple
Reynolds numbers.}
\label{fig:cp-curves}
\end{figure}

@@

The behavior of the power coefficient at $\lambda=3.1$ is plotted in
Fig.~\ref{fig:cp-re-dep}, where once again the increase in $C_P$ with $Re$ is
highlighted. We see that the power coefficient of the turbine increases
dramatically below $Re_D = 1 \times 10^6$ or $Re_c \approx 2 \times 10^5$,
beyond which there appears to be a small, linear, positive trend. The tendency
of $C_P$ to continue increasing slightly could be an effect of lower virtual
camber when compared to the strong $Re$-indepdendence of the higher $c/R$
UNH-RVAT \cite{Bachant2015-RVAT-Re-dep}.

@@

The mean velocity field at a tip speed ratio $\lambda=3.1$, 1 m downstream
($x/D=0.93$) is plotted in Fig.~\ref{fig:meancontquiv}. The mean streamwise
velocity deficit is markedly more symmetric than that of the UNH-RVAT, with
lower acceleration around the turbine thanks to lower rotor drag coefficient and
blockage ratio \cite{Bachant2015-JoT}. We also see that the effects of blade tip
vortex shedding are relatively weaker, which is likely an effect of the RM2's
lower solidity and tapered blades.

\begin{figure}[h]
\includegraphics[width=\textwidth]{figures/meancontquiv}
\caption{{\bf Mean Velocity.}
Mean velocity field at 1 m downstream ($x/D=0.93$), $U_\infty=1.0$ m/s, and
$\lambda=3.1$.}
\label{fig:meancontquiv}
\end{figure}

@@

Weighted averages for the transport terms in the mean kinetic energy
equation---analyzed identically to \cite{Bachant2015-JoT}---are shown in
Fig.~\ref{fig:Ktransport}. Due to the weaker blade vortex shedding, we see
transport due to vertical advection at this point in the wake is approximately 3
times lower than the higher solidity UNH-RVAT. Note that direct comparison is
somewhat invalid, since the total measurement plane area is about 5\% lower for
this experiment. However, the differences observed are beyond this discrepancy.
We also see relatively lower levels of cross-stream turbulent transport thanks
to the lack of blade stall vortex shedding. These results may have interesting
implications regarding the application of turbines with lower power coefficient
to possibly improve overall array performance through enhanced transport of
kinetic energy from the free stream, though a detailed anaylsis of the
downstream evolution and turbine interaction will be necessary.

@@

We see that strut drag losses increase with tip speed ratio to the power 2--3,
which makes streamlined struts much more important for low solidity turbines,
given the inverse correlation between solidity and $\lambda_0$.

>>>>

Sandia National Laboratories, in collaboration with the US Department of Energy
developed the Reference Model devices to provide modeling targets for advancing
marine hydrokinetic (MHK) energy technology \cite{Neary2014}. The Reference
Model 2 (RM2), a cross-flow turbine intended for river environments, was
initially designed by Barone et al. \cite{Barone2011} using the CACTUS vortex
line code \cite{Murray2011}. A scaled physical model of the RM2 was then
designed and built by the Saint Anthony Falls Laboratory (SAFL) to validate the
CACTUS predictions. Experimental performance measurements were significantly
lower than the numerical results \cite{Hill2014}, which was assumed to be a 
result of the physical model's low Reynolds number
\begin{equation}
Re_L = \frac{UL}{\nu},
\label{eq:Re}
\end{equation}\textbf{\textit{[why not go straight with defining ReD and ReC here, so it's also clear which U and L are used for each of these?]}}
where $U$ is ?free-stream? velocity, $L$ is characteristic length scales (?blade-chord length?), and $\nu$ is the fluid
kinematic viscosity.

As a cross-flow turbine, the RM2 blades experience large variations in angles of
attack as they rotate about their axis (``cross-flow turbine'' means the axis of
rotation is perpendicular to flow direction---either vertically or
horizontally). These variations become larger as the tip speed ratio
\begin{equation}
\lambda=\frac{\omega R}{U_\infty},
\end{equation}\textbf{\textit{[define Uinf. it looks like a list of notation at the end of the paper could help a alot]}}
where $\omega$ and $R$ are the rotor angular velocity and radius, decreases
\cite{Para2002}. Typically, the maximum angles of attack are sufficiently large
that the blades operate under dynamic stall, which is a complex unsteady process
and deviates significantly from static foil behavior, during part of the
rotation. Furthermore, the higher the solidity of a cross-flow turbine, the
lower the optimal tip-speed ratio at which it operates. Marine Hydrokinetic
(MHK) cross-flow turbines typically have higher solidity than cross-flow wind
turbines (VAWT), and hence operate at lower tip-speed ratios. Since MHK turbines
operate in a higher density fluid, unsteady dynamic effects related to the
blades' pitching motion and flow curvature also become more important when
compared to wind turbines.

@@

The need for experimental data that is relevant to full-scale behavior stems
from the need to validate numerical models---most importantly mid-fidelity
models---desirable for MHK developers to predict the performance of their
cross-flow turbine designs, since physical modeling at appropriate scales can be
prohibitively expensive in the early stages of engineering.

@@

A 1:6 scale model of the RM2 rotor was designed and built. Turbine geometric
parameters were scaled from the RM2 design report \cite{Barone2011},
with the exception of the shaft diameter, which was scaled from the SAFL RM2
model \cite{Hill2014}. The hubs were designed to be similar to the SAFL model,
which may aid in comparison of the results, though this was not a top priority.
Geometric parameters are shown in Tab.~\ref{tab:turb-geom} and a drawing of the
turbine design is shown in Fig.~\ref{fig:turbine-drawing}. The turbine model
components---blades, struts, shaft, and center hub sections---were fabricated
from 6061-T6 aluminum, which was hardcoat anodized per MIL-8625-A, type III,
class 2 specifications. CAD models of the turbine are available from
\cite{Bachant2015-RM2-CAD}.

@@

Experiments were performed in the UNH tow/wave tank, a 36 m long facility with a
3.66 m wide by 2.44 m deep cross-section, capable of tow speeds up to 3 m/s,
pictured in Figure~\ref{fig:tow-tank}\textbf{\textit{[missing fig?]}}. The turbine was mounted in a frame built
from NACA 0020 struts, attached to the tow carriage by four linear bearings,
which transfer all streamwise force to a pair of S-beam load cells \textbf{\textit{[need reference for each sensor - load cells, ADV, etc.]}}. The turbine
shaft RPM was controlled by a servo motor system, which allowed prescription of
the turbine tip speed ratio. The load torque was measured by an inline rotary
torque transducer. A load cell mounted, mounted at a fixed distance from the servo
motor, provides a redundant measurement. Turbine shaft angle was measured using
the servo drive's emulated encoder output, set to $10^5$ counts per turbine
shaft revolution. Carriage speed, and therefore inflow velocity was measured
using a linear encoder with 10 $\mu$m resolution. All of these
performance-related quantities were sampled at 2 kHz, while the tow tank's
motion controller provided redundant measurements of the carriage speed \textbf{\textit{[need to say how well the agreement between these two measurements are and which one is used throughout the paper; else, remove one of these]}} turbine angular velocity sampled at 1 kHz. Turbine wake measurements at 1 meter
downstream were measured with a Nortek Vectrino+ acoustic Doppler velocimeter,
sampling at 200 Hz. A photo and drawing of the experimental setup are shown in
Fig.~\ref{fig:exp-setup-photo} and Fig.~\ref{fig:exp-setup}, respectively.

@@

    \caption{{\bf Experimental setup.} Illustration of the experimental setup.}
    \textbf{\textit{[would be good to indicate the water surface with text of drawing ( e.g. a triangle and three lines]}}
    \label{fig:exp-setup}
\end{figure}

@@

\paragraph{Tare drag and torque compensation.} Tare torque and drag runs were
performed to measure the shaft bearing friction torque and turbine mounting
frame drag, respectively \textbf{\textit{[explain the purpose of these measurements; how the final drag and torque data were compensated using the tare measurements]}}. Tare drag runs were performed for each tow speed in
the experiment, for which the mean value was used in data processing\textbf{\textit{[describe how this was done - without blades but shaft?]}}. Tare
torque runs were performed by rotating the turbine shaft (without blades) in air
at constant angular velocity for a specified duration, over the range of angular
velocities used throughout the experiment. Tare torque was fit with a linear
regression versus shaft angular velocity, and added to the measured turbine
torque in post-processing.

@@

\textbf{\textit{[The list below is too detailed for a journal paper - suggest to remove the list and leave only the citation]}}
\begin{itemize}

@@

For each tow, a relevant quasi-steady duration was selected \textbf{\textit{[need a discussion on how this is done and justified]}}, and then truncated
to include a whole number of blade passages. Over this duration relevant
statistics are calculated.

@@

Mean rotor power and drag coefficients for multiple Reynolds numbers are plotted
versus tip speed ratio in Fig.~\ref{fig:cp-curves} and Fig.~\ref{fig:cd-curves},
respectively. $C_P$ is shown to increase in general with $Re$, along with a
reduction in the optimal tip speed ratio, due to stall delay \textbf{\textit{[need to explain what stall delay is]}}. These effect
diminish with increasing $Re$, which is expected as the blade boundary layers
transition from laminar to turbulent \textbf{\textit{[I wonder if this is relevant, since all cases are already in well in the turbulent regime]}} \cite{Lissaman1983,
Bachant2015-RVAT-Re-dep}. The rotor drag coefficient curves show similar
effects, but vary much less dramatically from each other.\textbf{\textit{[Overall I think we need more discussion for Figs 5 & 6.  For fig 5, maybe a discussion on stall delay and why Re independence is not achieve is sufficient.  Fig 6, need to find something to say, e.g. (1)why the trends look like that, is this expected? (2)Why they don't differ much between each other.    (3) mention that CD increases with ReD. ]}}

\begin{figure}[h]
\includegraphics[width=\textwidth]{figures/cp_curves}
\caption{{\bf Power Coefficient Curves.}
Mean rotor power coefficient plotted versus mean tip speed ratio for multiple
Reynolds numbers.}

\textbf{\textit{[This figure shows that we haven't achieved Re independence (close though). I am sure many others will notice this.  Need to state why we couldn't go over Re = 1.3E6. I presume due to technical limitation.]}}
\label{fig:cp-curves}
\end{figure}

@@

The behavior of the power coefficient at $\lambda=3.1$ is plotted in
Fig.~\ref{fig:cp-re-dep}, where once again the increase in $C_P$ with $Re$ is
highlighted. We see that the power coefficient of the turbine increases
dramatically below $Re_D = 1 \times 10^6$ or $Re_c \approx 2 \times 10^5$ \textbf{\textit{[define Rec]}},
beyond which there appears to be a small, linear, positive trend \textbf{\textit{[dCP/dRe seems to increase again for the two points: second and third highest Cp - any suggestion for the cause of this?  On a second thought, I think it's more appropriate to plot Re Vs. max Cp here ( don't fix TSR to 3.1, but pick a value which gives max Cp]}} . The tendency
of $C_P$ to continue increasing slightly could be an effect of lower virtual
camber \textbf{\textit{[I think we need to elaborate more on this. On its own, this sentence sounds very speculative]}} when compared to the strong $Re$-independence of the higher $c/R$
UNH-RVAT \cite{Bachant2015-RVAT-Re-dep}.

@@

The mean velocity field at a tip speed ratio $\lambda=3.1$, 1 m downstream
($x/D=0.93$) is plotted in Fig.~\ref{fig:meancontquiv}. The mean streamwise
velocity deficit is markedly more symmetric than that of the UNH-RVAT, with
lower acceleration around the turbine thanks to lower rotor drag coefficient and
blockage ratio \cite{Bachant2015-JoT}. We also see that the effects of blade tip
vortex shedding are relatively weaker, which is likely an effect of the RM2's
lower solidity and tapered blades.\textbf{\textit{[the last 2 sentences: it's difficult to understand this without knowing the UNH-RVAT. Suggest to add a discussion on the similarity & differences between RVAT and RM2 dimensions, testing, and add the mean velocity plot for RVAT side-by-side with the one for RM2.  Need more details on how you see the effects of vortex shedding are weaker.]}}

\begin{figure}[h]
\includegraphics[width=\textwidth]{figures/meancontquiv}
\caption{{\bf Mean Velocity.}
Mean velocity field at 1 m downstream ($x/D=0.93$), $U_\infty=1.0$ m/s, and
$\lambda=3.1$.}\textbf{\textit{[is this looking downstream or upstream? may want to indicate the rotation direction of the downstream rotor side ( right or left)]}}
\label{fig:meancontquiv}
\end{figure}

@@

Weighted averages for the transport terms in the mean kinetic energy
equation---analyzed identically to \cite{Bachant2015-JoT}\textbf{\textit{[Although this has been written in the cited reference, I feel we need to discuss this so that at least readers understand the reason of doing this analysis]}}---are shown in
Fig.~\ref{fig:Ktransport}. Due to the weaker blade vortex shedding, we see
transport due to vertical advection at this point in the wake is approximately 3
times lower than the higher solidity UNH-RVAT. Note that direct comparison is
somewhat invalid, since the total measurement plane area is about 5\% lower for
this experiment. However, the differences observed are beyond this discrepancy.
We also see relatively lower levels of cross-stream turbulent transport thanks
to the lack of blade stall vortex shedding. These results may have interesting
implications regarding the application of turbines with lower power coefficient
to possibly improve overall array performance through enhanced transport of
kinetic energy from the free stream, though a detailed anaylsis of the
downstream evolution and turbine interaction will be necessary.

@@

We see that strut drag losses increase with tip speed ratio to the power 2--3,
which makes streamlined struts much more important for low solidity turbines,
given the inverse correlation between solidity and $\lambda_0$.\textbf{\textit{[need more to write here]}}
